---
title: "Mass of chaffinches"
author: "Emma Rand"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
  bookdown::word_document2: default
bibliography: ../refs/reference.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.retina = 3)
options(digits = 2)
```

```{r packages}
library(tidyverse)
```

```{r functions}
source("sum_sq.R")
source("data_summary.R")
```

# Introduction {#introduction}

The chaffinch, *Fringilla coelebs*, is one of the most widespread and common passerine bird species [@cramp_1994]. It is sexually dimorphic in plumage with males being brightly coloured with a blue-grey cap and rust-red underparts and females being grey-brown with paler underparts (see Figure \@ref(fig:chaff-photo-fig). This study aimed to determine whether *F.coelebs* were also sexually dimorphic in size.

(ref:chaff-photo-fig) Adult *Fringilla coelebs* Left, male (by Andreas Trepte, CC BY-SA 2.5 <https://creativecommons.org/licenses/by-sa/2.5>, via Wikimedia Commons); right, female (4028mdk09, CC BY-SA 3.0 <https://creativecommons.org/licenses/by-sa/3.0>, via Wikimedia Commons)

```{r chaff-photo-fig, out.width="40%", fig.show='hold', fig.cap='(ref:chaff-photo-fig)'}
pics <- c("../pics/Chaffinch_(Fringilla_coelebs)male.jpg",
          "../pics/Fulda_Buchfinkweibchen_Juni_2012female.JPG")
knitr::include_graphics(pics)
```

# Methods

```{r import}
file <- "../data_raw/chaff.txt"
chaff <- read_table2(file)
```

```{r tidy}
chaff <- chaff %>%
  pivot_longer(names_to = "sex", 
               values_to = "mass",
               cols = everything())
```

```{r summary}
# sums of squares
sums_of_sq <- sum_sq(chaff, mass, sex)
# means, sd, se, n
chaff_sum <- data_summary(chaff, mass, sex)
```

```{r extract-summary}
# extract values for inline reporting
# ss(x)
m_ss <- sums_of_sq$ss[sums_of_sq$sex == "males"]
f_ss <- sums_of_sq$ss[sums_of_sq$sex == "females"]
# sample sizes
n_males <- chaff_sum$n[chaff_sum$sex == "males"] 
n_females <- chaff_sum$n[chaff_sum$sex == "females"] 
# highest mean, se and the sex to which they belongs
highestmean <- max(chaff_sum$mean) 
highestse <- chaff_sum$se[chaff_sum$mean == highestmean] 
highestsex <- chaff_sum$sex[chaff_sum$mean == highestmean]
# lowest mean, se and the sex to which they belongs
lowestmean <- min(chaff_sum$mean) 
lowestse <- chaff_sum$se[chaff_sum$mean == lowestmean] 
lowestsex <- chaff_sum$sex[chaff_sum$mean == lowestmean]
```

We randomly sampled `r n_males` male and `r n_females` female birds on the stray and determined their mass with spring scales. Analysis was carried out with R version 4.0.2 [@R-core] and tidyverse packages [@tidyverse].

# Results

```{r testing}
mod <- lm(data = chaff, mass ~ sex)
```

```{r extract-testing}
res <- summary(mod)
tval <- res$coefficients["sexmales", "t value"]
df <- res$df[2]
if (res$coefficients["sexmales", "Pr(>|t|)"] < 0.001) {
        b1p = "< 0.001"
        }
if (res$coefficients["sexmales", "Pr(>|t|)"] > 0.001) {
        b1p = paste("=", round(res$coefficients["sexmales", "Pr(>|t|)"], 3))
        }
rsq <- res$r.squared
```

We wouldn't normally report the sum of the squared deviations ($SS(x)$) but we do so here to make use of a function. For males, the $SS(x)$ = `r m_ss` and for females, the $SS(x)$ = `r f_ss`


`r str_to_sentence(highestsex)` ($\bar{x} \pm s.e.$: `r highestmean` $\pm$ `r highestse` g) have significantly higher mass than `r lowestsex` (`r lowestmean` $\pm$ `r lowestse` g)($t$ = `r tval`; $d.f.$ = `r df`; $p$ `r b1p`). See figure \@ref(fig:chaff-fig).

```{r chaff-fig, fig.height=4, fig.width=4, fig.cap="(ref:chaff-fig)"}
# note that the axis limits and significance labelling are not very reproducible here
# they would likely need updating if the data altered.
ggplot() +
  geom_jitter(data = chaff, 
              aes(x = sex, y = mass), 
              width = 0.25, colour = "grey") +
  geom_errorbar(data = chaff_sum,
                aes(x = sex,
                    ymin = mean,
                    ymax = mean),
                width = .3) +
  geom_errorbar(data = chaff_sum,
                aes(x = sex,
                    ymin = mean - se,
                    ymax = mean + se),
                width = .5) +
  geom_segment(aes(x = 1, y = 28, xend = 2, yend = 28),
               size = 1) +
  geom_segment(aes(x = 1, y = 28, xend = 1, yend = 27),
               size = 1) +
  geom_segment(aes(x = 2, y = 28, xend = 2, yend = 27),
               size = 1) +
  annotate("text", x = 1.5, y = 29,  label = paste("p",b1p)) +
  scale_x_discrete(labels = c("Females", "Males"),
                   name = "") +
  scale_y_continuous(name = "Mass (g)",
                     expand = c(0, 0),
                     limits = c(0, 30)) +
  theme_classic()
```
(ref:chaff-fig) Mass of adult chaffinches caught on the Stray. Error bars are $\bar{x} \pm 1 s.e.$   

I've also gratuitously included a table with the same information just for the sake of including a table. See Table \@ref(tab:summary-table).

```{r summary-table}
knitr::kable(chaff_sum,
             digits = 4,
             caption = 'A summary of the data.',
             row.names = FALSE)
```


# Discussion

Here we pick up points from the introduction. See \@ref(introduction).

# References